Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-04-14 paper/intro.tex
update macros to distinguish REL and ACQ
intro

This paper is about the interaction of two of the fundamental building
blocks of computing: memory and sequential composition. One would like
to think that these are well-worn topics, where every issue has been
settled, but this is sadly not the case.

\subsection{Memory}

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order of reads and writes,
where each read has a matching \emph{fulfilling} write,
for example:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2\SEMI
    r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx1}{\DR{x}{1}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=20,in=160]{wx1}{rx1}
        \po{wx0}{wx1}
        \po{wx1}{wy2}
        \po{wy2}{ry2}
        \po{ry2}{rx1}
      \end{tikzinline}}
  \end{gather*}
(In examples, $\aReg$--$\bReg$ range over thread-local registers and $\aLoc$-$\cLoc$
range over shared memory references.)

This model naturally extends to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
    \PAR
    \THREAD{r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx1}{\DR{x}{1}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=20,in=160]{wx1}{rx1}
        \po{wx0}{wx1}
        \po{wx1}{wy2}
        \po{ry2}{rx1}
      \end{tikzinline}}
  \end{gather*}

Unfortunately, this model does not compile efficiently to commodity
hardware, resulting in a 37--73\% increase in CPU time on ARM~\cite{Liu:2019:ASC:3314221.3314611} and,
hence, in power consumption.  Developers of software and compilers have
therefore been faced with a difficult trade-off, between an elegant
model of memory, and its impact on resource usage (such as size of
data centers, electricity bills and carbon footprint). Unsurprisingly,
many have chosen to prioritize efficiency over elegance.

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location} and not globally
over the whole program. This allows executions which
are inconsistent with program order, such as:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
    \PAR
    \THREAD{r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx0}{\DR{x}{0}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=15,in=165]{wx0}{rx0}
        \po{wx0}{wx1}
        \po[out=200,in=340]{rx0}{wx1}
      \end{tikzinline}}
  \end{gather*}

In such models, the causal order between events is important,
and includes control and data dependencies, to avoid
paradoxical ``out of thin air'' examples such as:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \FI}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI y\GETS1}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}

Unfortunately, while a simple syntactic approach to dependency calculation
suffices for hardware models, it is not preserved by common compiler
optimizations. For example, if we calculate control dependencies syntactically,
then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \ELSE y\GETS1 \FI}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
An optimizing compiler might lift the assignment $y\GETS1$ out of the conditional,
thus removing the control dependency.

\end{itemize}

In this paper, we will focus on logical models, as those are compositional,
and align well with existing models of sequential composition.
The heart of the model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20} is to add logical preconditions
to events, which are introduced by store actions (modeling data dependencies)
and conditionals (modeling control dependencies):
  \begin{gather*}
    \IF{s{<}1} \THEN z\GETS r{*}s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
      \end{tikzinline}}
  \end{gather*}
Preconditions are discharged by being ordered after a read:
  \begin{gather*}
    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
        \po{ry0}{wz0}
      \end{tikzinline}}
  \end{gather*}
Note that there is dependency order from $(\DR{y}{0})$ to $(\DW{z}{0})$
so the precondition for $(\DW{z}{0})$ only has to be satisfied assuming the hypothesis
$(s{=}0)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
which is why we do not assume the hypothesis $(r{=}0)$. Nonetheless, the precondition on
$(\DW{z}{0})$ is a tautology, and so can be elided in the diagram:
  \begin{gather*}
    % r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
    % \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \event{wz0}{\DW{z}{0}}{right=of ry0}
        \po{ry0}{wz0}
      \end{tikzinline}}
  \end{gather*}

While existing models of relaxed memory have detailed treatments of parallel composition,
they often give sequential composition little attention, either ignoring it altogether,
or treating it operationally with its usual small-step semantics. This paper
investigates how existing models of sequential composition interact with relaxed memory.

\subsection{Sequential composition}

%% Program logics and semantics tell us that when executing ((S1; S2), state0),
%% we execute (S1, state0) to arrive at state1, then execute (S2, state1) to
%% arrive at the final state2.

%% This is a delightfully simple story that can be explained to children.  It is
%% also a lie.

%% Processors execute instructions out of order, due to pipelines and caches.
%% Compilers reorder programs even more dramatically.  All of this reordering is
%% meant to be unobservable in single-threaded code.  In multi-threaded code,
%% however, all bets are off.  A formal attempt to understand the resulting mess
%% is known as a ``relaxed memory model.''

%% Most of the relaxed memory models that have been proposed are designed to
%% help us understand whole program execution: they have no compositionality
%% properties whatsoever.  Recently, denotation models have appeared that treat
%% \emph{concurrent} execution compositionally.  One such model is ``Pomsets with
%% Preconditions''.  It remains an open question, however, whether it is
%% possible to treat \emph{sequential} execution compositionally in such a model,
%% without overly restricting processors and compilers.

%% We propose adding families of predicate transformers to Pomsets with
%% Preconditions.  The resulting model is denotational, supporting both parallel
%% and sequential composition.  When composing (S1;S2), the predicate
%% transformer used to validate the precondition of an event in S2 is chosen
%% based on the dependency order from S1 into this event.  As usual in work on
%% relaxed memory, we have not handled loops or recursion.

%% Happily, most of the results expected of a relaxed memory model can be
%% established by appeal to prior work.  So here we are able to concentrate on
%% the model itself.  The model is formalized in Agda, where we have established
%% associativity for sequential composition.

%% For the memory-model specialist, we retain the good properties of the prior
%% work on Pomsets with Preconditions, fixing some errors along the way.  These
%% properties include efficient implementation on ARMv8, support for compiler
%% optimizations, support for logics that prove the absence of thin-air
%% behaviors, and a local data race freedom theorem.

@@

Our approach follows that of weakest precondition semantics of
\citet{DBLP:journals/cacm/Dijkstra75}, which provides an alternative
characterization of Hoare logic \citep{Hoare:1969:ABC:363235.363259} by
mapping postconditions to preconditions. We recall the definition of
$\fwp{\aCmd}{\bForm}$ for loop-free code below. % in Figure~\ref{fig:sp}
\begin{itemize}
\item
  \begin{math}
    \fwp{\SKIP}{\bForm} = \bForm
  \end{math}
\item 
  \begin{math}
    \fwp{\ABORT}{\bForm} = \FALSE
  \end{math}
\item
  \begin{math}
    \fwp{r\GETS \aExp}{\bForm} = \bForm[\aExp/r]
  \end{math}
\item
  \begin{math}
    \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
  \end{math}
\item
  \begin{math}
    \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
  \end{math}
  \\
  \begin{math}
    ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
    ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
  \end{math}
\end{itemize}
The rule we are most interested
in is the one for sequential composition, which maps sequential composition of programs
to function composition of predicate transformers.

Predicate transformers are a good fit to logical models of dependency calculation,
since both are concerned with preconditions, and how they are transformed by
sequential composition. Our first attempt is to associate a predicate transformer
with each pomset. We visualize this in diagrams by showing how $\bForm$ is transformed,
for example:
  \begin{align*}
    \begin{gathered}
      r \GETS x
      \\
      \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
          \event{rx0}{\DR{x}{0}}{}
          \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
          \xo{rx0}{rx0d}
        \end{tikzinline}}
    \end{gathered}
    &&
    \begin{gathered}
      s \GETS y
      \\
      \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
          \event{ry0}{\DR{y}{0}}{}
          \xform{ry0d}{(s{=}0) \limplies \bForm}{below=of ry0}
          \xo{ry0}{ry0d}
        \end{tikzinline}}
    \end{gathered}
    &&
    \begin{gathered}
      \IF{s{<}1} \THEN z\GETS r{*}s \FI
      \\
      \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
          \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
          \xform{wz0d}{\bForm}{below=of wz0}
          \xo[out=-30]{wz0}{wz0d}
      \end{tikzinline}}
    \end{gathered}
  \end{align*}
In the rightmost program above, the write to $z$ affects the shared store, not the
local state of the thread, therefore we assign it the identity transformer.

>>>>

This paper is about the interaction of two of the fundamental building blocks
of computing: sequential composition and mutable state. One would like to
think that these are well-worn topics, where every issue has been settled,
but this is not the case.

\subsection{Sequential Composition} %Predicate Transformers}

Introductory programmers are taught \emph{sequential abstraction}: that the
program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
the late 60s, we've been able to explain this using logic
\citep{Hoare:1969:ABC:363235.363259}.  In
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s
[\citeyear{DBLP:journals/cacm/Dijkstra75}] formulation, we think of programs
as \emph{predicate transformers}, where predicates describe the state of
memory in the system.  In the calculus of weakest preconditions, programs map
postconditions to preconditions.  We recall the definition of
$\fwp{\aCmd}{\bForm}$ for loop-free code below. 
\begin{multicols}{2}
  \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
  \item
    \begin{math}
      \fwp{\SKIP}{\bForm} = \bForm
    \end{math}
    % \item 
    %   \begin{math}
    %     \fwp{\ABORT}{\bForm} = \FALSE
    %   \end{math}
    \stepcounter{enumi}
  \item[] \labeltext[\textsc{d}2]{}{wp-assign}
    \begin{enumerate}[leftmargin=0pt]
    \item \label{wp-write}
      $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
    \item \label{wp-let}
      $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
    \item \label{wp-read}
      $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
    \end{enumerate}
    % \item 
    %   \begin{math}
    %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
    %   \end{math}
  \item
    \begin{math}
      \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
    \end{math}
  \item
    \begin{math}
      \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
    \end{math}
    \\
    \begin{math}
      ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
      ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
    \end{math}
  \end{enumerate}
\end{multicols}
For this language, the Hoare triple
$\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
\fwp{\aCmd}{\bForm}$.  We have split
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
\eqref{wp-assign} into three cases. In our notation,
$\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
most once,
$\aLoc$--$\cLoc$ range over shared memory references, and
$\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
include
$\aLoc$--$\cLoc$.\footnote{We wrote \ref{wp-read} using implication rather
  than substitution; the two formulations are equivalent under the assumption
  that each register is assigned at most once.}

This is quite a pretty explanation of sequential computation in a sequential
context.  In a concurrent context, however, \ref{wp-read} is unsound!
\ref{wp-read} assumes that a read from
$x$ must be fulfilled by a preceding write to
$x$.  In a concurrent context, writes may come from other threads.

Existing approaches to sequential composition in the concurrent context
either assume exclusive access, as in concurrent separation logic
\cite{OHearn:2007:RCL:1235896.1236121}, or abandon the logical approach
altogether, as in the pomset model of
\citet{DBLP:journals/corr/abs-1804-04214}---this model uses syntactic
dependencies and thus dramatically limits compiler optimization.  This leaves
open the question of how to apply logic to programs with data races.
To understand the solution, one must first understand the constraints imposed
by hardware and compilers.

% Assuming all variables are initialized to $0$,
% $\PW{x}{1}\SEMI\PW{y}{1}$ can be distinguished from $\PW{y}{1}\SEMI\PW{x}{1}$
% by the concurrent observer $\PR{x}{r}\SEMI\PR{y}{s}$

% Microprocessors and compilers do not execute this way, but they attempt to
% preserve the abstraction, at least for sequential code.  For concurrent code,
% all bets are off.

% \citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
% semicolon'' is problematic for programmers, and thus concurrent languages
% should guarantee sequential abstraction.  But processor and language
% implementors balk at the prospect, due to significant costs.

% These leaves programmers in a bit of pickle: What does that semicolon mean?

% The last decade has seen a slew of research in ``relaxed memory models,''
% which attempt to explain concurrent execution.  The models are mostly
% operational, sometimes with an additional series of axioms to eliminate
% ``bad'' executions.  These models have limited modularity properties,
% typically requiring whole program analysis.

% In this paper, we attempt to

\subsection{Memory Models}

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order of reads and writes,
where each read has a matching \emph{fulfilling} write,
for example:
  \begin{gather*}
    \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}\SEMI
    \PR{y}{r}\SEMI \PR{x}{s}}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=3em of wy2}
        \event{rx1}{\DR{x}{1}}{right=of ry2}
        \rf[out=15,in=165]{wy2}{ry2}
        \rf[out=15,in=165]{wx1}{rx1}
        \po{wx0}{wx1}
        \po{wx1}{wy2}
        \po{wy2}{ry2}
        \po{ry2}{rx1}
      \end{tikzinline}}
  \end{gather*}
% (In examples, $\aReg$--$\bReg$ range over thread-local registers and $\aLoc$-$\cLoc$
% range over shared memory references.)

This model naturally extends to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example:
  \begin{gather*}
    \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
    \PAR
    \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=3em of wy2}
        \event{rx1}{\DR{x}{1}}{right=of ry2}
        \rf[out=15,in=165]{wy2}{ry2}
        \rf[out=15,in=165]{wx1}{rx1}
        \po{wx0}{wx1}
        \po{wx1}{wy2}
        \po{ry2}{rx1}
      \end{tikzinline}}
  \end{gather*}

Unfortunately, this model does not compile efficiently to commodity
hardware, resulting in a 37--73\% increase in CPU time on Arm8~\cite{Liu:2019:ASC:3314221.3314611} and,
hence, in power consumption.  Developers of software and compilers have
therefore been faced with a difficult trade-off, between an elegant
model of memory, and its impact on resource usage (such as size of
data centers, electricity bills and carbon footprint). Unsurprisingly,
many have chosen to prioritize efficiency over elegance.

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location} and not globally
over the whole program. This allows executions which
are inconsistent with program order, such as:
  \begin{gather*}
    \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
    \PAR
    \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=3em of wy2}
        \event{rx0}{\DR{x}{0}}{right=of ry2}
        \rf[out=15,in=165]{wy2}{ry2}
        \rf[out=15,in=165]{wx0}{rx0}
        \po{wx0}{wx1}
        \po[out=-165,in=-15]{rx0}{wx1}
      \end{tikzinline}}
  \end{gather*}

In such models, the causal order between events is important,
and includes control and data dependencies, to avoid
paradoxical ``out of thin air'' examples such as:
  \begin{gather*}
    \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \FI}
    \PAR
    \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=3em of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=15,in=165]{wy1}{ry1}
        \rf[out=165,in=15]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
  \begin{gather*}
    \THREAD{\PR{x}{r}\SEMI \PW{y}{1}}
    \PAR
    \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=3em of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=15,in=165]{wy1}{ry1}
        \rf[out=165,in=15]{wx1}{rx1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}

Unfortunately, while a simple syntactic approach to dependency calculation
suffices for hardware models, it is not preserved by common compiler
optimizations. For example, if we calculate control dependencies syntactically,
then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
  \begin{gather*}
    \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \ELSE \PW{y}{1} \FI}
    \PAR
    \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=3em of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=15,in=165]{wy1}{ry1}
        \rf[out=165,in=15]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
An optimizing compiler might lift the assignment $\PW{y}{1}$ out of the conditional,
thus removing the control dependency.

To address this, \citet{DBLP:journals/pacmpl/JagadeesanJR20} introduced
\emph{Pomsets with Preconditions}, where events are labeled with logical
formulae.  Nontrivial preconditions are introduced by store actions (modeling
data dependencies) and conditionals (modeling control dependencies):
  \begin{gather*}
    \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
      \end{tikzinline}}
  \end{gather*}
Preconditions are discharged by being ordered after a read:
  \begin{gather*}
    \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \event{wz0}{(0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
        \po{ry0}{wz0}
      \end{tikzinline}}
  \end{gather*}
Note that there is dependency order from $(\DR{y}{0})$ to $(\DW{z}{0})$
so the precondition for $(\DW{z}{0})$ only has to be satisfied assuming the hypothesis
$(0{=}s)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
which is why we do not assume the hypothesis $(0{=}r)$. Nonetheless, the precondition on
$(\DW{z}{0})$ is a tautology, and so can be elided in the diagram:
\begin{gather*}
  % \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  % \\
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx0}{\DR{x}{0}}{}
      \event{ry0}{\DR{y}{0}}{right=of rx0}
      \event{wz0}{\DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
    \end{tikzinline}}
\end{gather*}

\subsection{Predicate Transformers For Relaxed Memory}

Pomsets with Preconditions show how the logical approach to sequential
dependency calculation can be mixed into a relaxed memory model.  However,
\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} do not provide a model of
sequential composition.  Instead, their model uses \emph{prefixing}, which
requires that the model is built from right to left: events are prepended one
at a time, with perfect knowledge of the future.  This makes reasoning about
sequential program fragments difficult.  For example,
\citeauthor{DBLP:journals/pacmpl/JagadeesanJR20} state the equivalence
allowing reordering independent writes as follows,
\begin{align*}
  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp\SEMI\aCmd} &=
  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp\SEMI\aCmd} \;\;\text{if}\; \aLoc\neq\bLoc
\end{align*}
where $\aCmd$ is the entire future computation!  By formalizing sequential
composition, we can show:
\begin{align*}
  \sem{\aLoc \GETS \aExp \SEMI \bLoc  \GETS \bExp} &=
  \sem{\bLoc  \GETS \bExp\SEMI \aLoc \GETS \aExp} \;\;\text{if}\; \aLoc\neq\bLoc
\end{align*}
Then the equivalence holds in any (sequential) context.

% While existing models of relaxed memory have detailed treatments of parallel composition,
% they often give sequential composition little attention, either ignoring it altogether,
% or treating it operationally with its usual small-step semantics. This paper
% investigates how existing models of sequential composition interact with relaxed memory.

@@

% Our approach follows that of weakest precondition semantics of
% \citet{DBLP:journals/cacm/Dijkstra75}, which provides an alternative
% characterization of Hoare logic \citep{Hoare:1969:ABC:363235.363259} by
% mapping postconditions to preconditions. We recall the definition of
% $\fwp{\aCmd}{\bForm}$ for loop-free code below. % in Figure~\ref{fig:sp}
% \begin{itemize}
% \item
%   \begin{math}
%     \fwp{\SKIP}{\bForm} = \bForm
%   \end{math}
% % \item 
% %   \begin{math}
% %     \fwp{\ABORT}{\bForm} = \FALSE
% %   \end{math}
% \item
%   \begin{math}
%     \fwp{\LET{r}{\aExp}}{\bForm} = \bForm[\aExp/r]
%   \end{math}
% \item
%   \begin{math}
%     \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
%   \end{math}
% \item
%   \begin{math}
%     \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
%   \end{math}
%   \\
%   \begin{math}
%     ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
%     ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
%   \end{math}
% \end{itemize}
% The rule we are most interested
% in is the one for sequential composition, which maps sequential composition of programs
% to function composition of predicate transformers.

Predicate transformers are a good fit to logical models of dependency calculation,
since both are concerned with preconditions, and how they are transformed by
sequential composition. Our first attempt is to associate a predicate transformer
with each pomset. We visualize this in diagrams by showing how $\bForm$ is transformed,
for example:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of rx0}
        \xo{rx0}{rx0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \PR{y}{s}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{ry0}{\DR{y}{0}}{}
        \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
        \xo{ry0}{ry0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
        \xform{wz0d}{\bForm[r{*}s/z]}{right=of wz0}
        \xo{wz0}{wz0d}
      \end{tikzinline}}
  \end{gathered}
\end{align*}
The predicate transformer from the write matches
\citeauthor{DBLP:journals/cacm/Dijkstra75}.  For the reads, however,
\ref{wp-read} defines the transformer of $\PR{x}{r}$ to be
$(x{=}r) \limplies \bForm$.  Instead, we use $(0{=}r) \limplies \bForm$,
reflecting the fact that $0$ may come from a concurrent write.  The
obligation to find a matching write is moved from the sequential semantics of
\emph{substitution} and \emph{implication} to the concurrent semantics of
\emph{fulfillment}.

% In the rightmost program above, the write to $z$ affects the shared store, not the
% local state of the thread, therefore we assign it the identity transformer.

For the sequentially consistent semantics, sequential composition is
straightforward: we apply each predicate transformer to the preconditions of
subsequent events, composing the predicate transformers.  (In subsequent
diagrams, we only show predicate transformers for reads.)
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  \\[-1.5ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{rx0}{\DR{x}{0}}{}
      \event{ry0}{\DR{y}{0}}{right=of rx0}
      \event{wz0}{(0{=}r) \limplies (0{=}s) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
      \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{left=of rx0}
      \po{rx0}{ry0}
      \po{ry0}{wz0}
      \xo{rx0}{rx0ry0d}
      \xo[out=155,in=15]{ry0}{rx0ry0d}
      % \xo{wz0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
This model works for the sequentially consistent case, but needs to be
weakened for the relaxed case. The key observation of this paper is
that rather than working with one predicate transformer, we should
work with a \emph{family} of predicate transformers, indexed by sets
of events.
